---
title: "Humdrum"
subtitle: "A package designed to remove the humdrum from Exploratory Data Analysis using the AvoNet dataset"
author: "Charlotte Drury U6603688"
date: "04/10/2023"
output:  
    html_document:
        toc: true
        toc_depth: 4
        theme: cosmo
        number_sections: false
        toc_float: true
        highlight: pygments
        fig_width: 8
        fig_height: 4
---

# Word/figure count

words: 1994

figures: 0

# Location on GitHub

https://github.com/DruryCF/DS4B-final-project

# Data Description

The AvoNet dataset contains functional trait data for 90,020 individual birds representing 11,009 extant species sampled from 181 countries (Tobias et al., 2022). It includes 11 continuous morphological traits (such as beak and wing dimensions; and tail, tarsus, and body size), 6 ecological variables, as well as geographic data. AvoNet also includes species averages according to three different taxonomy classifications to enable integration with other databases.

# Questions/Aims

Aim: to build an R-package (`humdrum`) to interface with AvoNet

* Build functions to:
    + install and load AvoNet as a relational database (stored as a list of tibbles)
    + tidy data when loading (largely correcting variable class and renaming as AvoNet is already in tidy format)
    + clean data when loading (toggled based on user preference)
    + facilitate Exploratory Data Analysis (EDA) (extract defined variables according to different taxonomies and apply summary statistics)

Justification: 

AvoNet is difficult to explore and interact with as it is stored across 11 related excel spreadsheets. Other datasets stored similarly (e.g. AusTraits, Atlas of Living Australia) have dedicated R-packages to streamline their use (Falster et al., 2021; Westgate et al., 2023). Some of AvoNet is incorporated in the `traitdata` package, however this (at the time of writing) is only a single observed individual per species according to the BirdLife taxonomy classification (RS-eco, 2022). An R-package that interfaces with the entirety of AvoNet would facilitate EDA and potentially aid in addressing specific research questions.

# Raw data

The raw data is obtainable from: https://figshare.com/ndownloader/files/34480856?private_link=b990722d72a26b5bfead; linked as `AVONET Supplementary dataset 1` in Tobias and colleagues (2022).

`humdrum`'s function, `avonet_install()`, accomplishes downloading AvoNet for the user. `avonet_install()` only takes one argument, the file destination of the downloaded file (`filename`).

```{r}
#devtools::install_github("DruryCF/DS4B-final-project", subdir = "humdrum")
library(humdrum)
library(readxl)
library(tidyverse)
library(skimr)
```

```{r}
avonet_install(filename = "raw_data/avonet_dataset.xlsx")
```

Should additional versions of AvoNet be released in the future and older versions remain available via separate link, a version argument may be added to `avonet_install()` to enable users to access specific releases of AvoNet.

# Data wrangling

`humdrum`'s function `avonet_load()` automatically tidies and optionally cleans the AvoNet dataset. Currently, cleaning is limited to the `raw_data` tibble as the `birdtree`, `ebird`, and `birdlife` tibbles were determined to not require cleaning (for details see `processed_data/data_cleaning.Rmd`). The `clean` argument can variably clean some of the data based on defined criteria (e.g. `clean = "gapspecies"` will remove and relocate all species with no observations from the `raw_data` tibble to a new `gapspecies` tibble). When cleaning is applied, the `avonet_exclude` object will be written to the global environment, detailing removed observations. The `excluded_data_raw_data.csv` found in `/processed_data` and produced in `data_cleaning.Rmd` details all cleaning options (apart from those concerning variables rather than observations) and how to execute them through the `clean` argument.

```{r}
avonet_tidy <- avonet_load(filename = "raw_data/avonet_dataset.xlsx")
avonet_clean <- avonet_load(filename = "raw_data/avonet_dataset.xlsx", clean = "all")

avonet_tidy$raw_data
avonet_clean$raw_data
avonet_clean$gapspecies
avonet_exclude$raw_data
```

`avonet_load()` uses `read_xlxs()` from the `readxl` package (under the `tidyverse` package v2.0.0) to read each excel sheet in the AvoNet dataset. It uses the `colnames()` and `tolower()` functions to make all variable names lowercase to increase consistency and ease of use (additionally renames `hand-wing.index` with `hand_wing.index`). It also corrects variable class where necessary (usually numeric or character to factor). For example:

```{r}
raw_data <- read_xlsx("raw_data/avonet_dataset.xlsx", sheet = "AVONET_Raw_Data", na = "NA")

# Add row index variable; This is not included in avonet_load() but is used to subtract excluded observations from tidied_data_*.csv. 

raw_data <- raw_data %>%
  mutate(row_index = row_number(),
         .before = 0)

# Make variables lowercase across tibbles and rename hand-wing.index

colnames(raw_data) <- tolower(colnames(raw_data))

raw_data <- raw_data %>% 
  rename(hand_wing.index = `hand-wing.index`)

# correct variable class

raw_data$data.type <- raw_data$data.type %>%  
  as.factor()

raw_data$protocol <- raw_data$protocol %>% 
  as.factor()

raw_data$age <- raw_data$age %>% 
  as.factor()

raw_data$sex <- raw_data$sex %>% 
  as.factor()

# write .csv

write.csv(raw_data, file = "processed_data/tidied_data_raw_data.csv")
```

This tidied version of the `raw_data` tibble in AvoNet can be found as a separate `.csv` in `/processed_data` along with the tided versions of the `birdlife`, `ebird`, and `birdtree`, tibbles. The code behind these files can be found in `data_cleaning.Rmd`.

While `avonet_load()` can optionally clean the data, the `excluded_data_raw_data.csv` can also be used to remove the relevant observations from `tidied_data_raw_data.csv`. Note that this will not remove the `locality` and `country` variables which are removed in `avonet_load()` by setting clean to `"empty"` or `"all"` as the data for these variables is currently embargoed.

```{r}
raw_data_tidy <- read_csv("processed_data/tidied_data_raw_data.csv")
raw_data_exclude <- read_csv("processed_data/excluded_data_raw_data.csv")

raw_data_clean <- raw_data_tidy %>% 
  filter(!(row_index %in% raw_data_exclude$row_index))

write.csv(raw_data_clean, file = "processed_data/cleaned_data_raw_data.csv")

raw_data_clean
raw_data_exclude
```

`avonet_load()` also adds an additional tibble to the original dataset under `[object_name]$description` that describes each tibble or excel spreadsheet in the original dataset (Note that it does not describe the `gapspecies` tibble created using the `clean` argument).

```{r}
avonet_clean$description
```


# Sanity checks

To check that the raw data is clean, we can use `skim()` from the `skimr` package.

```{r}
skim(avonet_clean$raw_data)
```

The data has the expected number of continuous morphological traits (10), the 11th morphological trait, mass, is stored exclusively in the interspecies tibbles (i.e. `birdlife`, `ebird`, `birdtree`; these exclude intraspecies variation as values are species means) as the trait was extracted from literature for each species and not measured for individuals. The data has the expected number of rows (19,019; from 19,020 individuals where one removed as unidentified species - this individual can be optionally included by instructing the `clean` argument in `avonet_load()`). The range across the numeric traits seems reasonable, given knowledge of variation between bird species.

# Addressing the questions/aims

Since it was published last year, the original paper releasing AvoNet (Tobias et al., 2022) has been cited at least 115 times. From a sample of 25 citing papers 7 used the AvoNet dataset as a major contributor (using several variables; e.g. Reid et al., 2023) while 11 used the dataset in a minor way (using single values or variables; e.g. Lin et al., 2023). `humdrum` is an R package (dependent on the `tidyverse` package (built under v2.0.0) including non-core package `readxl`) designed to interface with AvoNet. `humdrum` is intended to ease exploratory data analysis (EDA) and answering research questions using AvoNet. Furthermore, interacting with and using AvoNet through `humdrum` can increase the reproducibility of research like similar packages aim to do (e.g. Román-Palacios, 2023; Westgate et al., 2023). Reproducibility is a major issue across sciences including ecology and evolution (Parker et al., 2016). Methodology (including cleaning, editing, code, etc.) and results are often reported in insufficient detail to enable the reproduction of research. Thus, interpretation of and building on this research is impaired significantly.

Reid and colleagues (2023) used AvoNet variables to calculate a wing loading index as a proxy for flight manoeuvrablity. They calculated this variable as body mass / (wing length * wing width). While body mass corresponds to the `mass` variable in AvoNet and wing length presumably corresponds to `wing.length` (although AvoNet includes additional measures of wing length), the variable they use for wing width is uncertain. Most likely, `secondary1` was used as it measures approximately perpendicular to `wing.length` in many species (Tobias et al., 2023) Additionally, it is uncertain which taxonomy classification they performed their calculations using. This is an example of under-reporting of methodology which hampers reproducibility; although the data is publicly available, it is uncertain what Reid and colleagues did to achieve their results. Using `humdrum` to interface with AvoNet and `tidyverse` functions to perform calculations, Reid and colleagues could have easily built an entirely reproducible workflow.

Assuming Reid and colleagues used the `mass`, `wing.length`, and `secondary1` variables according to the birdlife taxonomy to calculate their wing loading index (FM), `humdrum` and `tidyverse` can be used to reproduce their calculations.

After using `avonet_install()`, `avonet_extract_interspecies()` can be used to bypass `avonet_load()` (as it calls this function directly) and select the needed variables (along with other relevant information).

```{r}
reid_variables <- avonet_extract_interspecies(filename = "raw_data/avonet_dataset.xlsx",
                                              taxon_system = "birdlife",
                                              traits = c("mass", "wing.length", "secondary1"))
reid_variables
```

Using `mutate()` from `tidyverse`, the raw wing loading index is quickly calculated.

```{r}
reid_variables <- reid_variables %>% 
  mutate(species = species,
         FM_raw = mass / (wing.length * secondary1),
         .keep = "used")

reid_variables
```

Again with `mutate()`, the index can be reformatted as a scale of 1-5 as in Reid and colleagues original work.

```{r}
reid_variables <- reid_variables %>%
  mutate(species = species,
         FM = case_when(FM_raw < 0.01 ~ 1,
                        FM_raw >= 0.01 & FM_raw < 0.02 ~ 2,
                        FM_raw >= 0.02 & FM_raw < 0.03 ~ 3,
                        FM_raw >= 0.03 & FM_raw < 0.04 ~ 4,
                        FM_raw > 0.04 ~ 5),
         .keep = "none") %>% 
  arrange(desc(FM))

reid_variables
```

Although some methodology was assumed, these values appear to match Reid and colleagues', highlighting how `humdrum` could aid in reproducibility.

AvoNet's interspecies tibbles (i.e. `birdlife`, `ebird`, `birdtree`) provide the means of species for AvoNet's functional traits including those only available at species level such as geographic information (and `mass` as it was extracted from literature). This means that no intraspecies variation is included in this data, unlike the `raw_data` tibble. Additionally, the interspecies tibbles record observations as sex general meaning any sex specific information and variation is lost. `humdrum`'s `avonet_extract_intraspecies()` function enables extraction of traits (and other variables) from the `raw_data` tibble of AvoNet. Furthermore, it applies the specified taxonomy classification system to the data. This enables exploration and use of what intraspecies variation is available in AvoNet including in analysis between sexes and higher taxonomic levels.

```{r}
intraspecies <- avonet_extract_intraspecies(filename = "raw_data/avonet_dataset.xlsx",
                                            taxon_system = "ebird",
                                            traits = c("sex", "mass", "wing.length", "secondary1"),
                                            clean = "all")
intraspecies
```

Using `humdrum`'s `avonet_summary()` function (which can also be applied to the output of `avonet_extract_interspecies()`), we can apply summary statistics to this extracted data.

```{r}
intraspecies_summary <- avonet_summary(data = intraspecies, group = c("species", "sex"), stats = c("mean", "median", "sd"), tidy = FALSE)
intraspecies_summary
```

This function defaults to a non-tidy format as it is easier to explore visually. For analysis, the `tidy` argument can be set to `TRUE`.

```{r}
intraspecies_summary <- avonet_summary(data = intraspecies, group = c("species", "sex"), stats = c("mean", "median", "sd"), tidy = TRUE)
intraspecies_summary
```

Using the means of `wing.length` and `secondary1`, we can almost repeat the calculation for wing loading index on sex specific data. Unfortunately, AvoNet only includes body mass data at a species, sex general level. Therefore, an outside mass source would be needed.

It is important to note that AvoNet includes a very limited number of individuals per species as it focuses on broad species coverage. Therefore, AvoNet is limited in answering particular research questions. However, it is perfectly built for exploration. Using `humdrum` to interface with AvoNet, researchers can easily navigate and explore while making their work reproducible. `humdrum` takes care of the bulk of the data retrieval and manipulation. Thus, used in conjunction with visualisation packages like `ggplot2`, it could be a powerful tool for reproducible EDA. Furthermore, as AvoNet expands and updates, `humdrum` and AvoNet could become additionally useful for answering particular research questions.

# References

Falster, D., Gallagher, R., Wenk, E.H. et al. (2021). AusTraits, a curated plant trait database for the Australian flora. _Sci Data, 8_, 254. https://doi.org/10.1038/s41597-021-01006-6

Lin, Y., Chen, Y., Rubenstein, D.R. et al. (2023). Environmental quality mediates the ecological dominance of cooperatively breeding birds. _Ecology Letters, 26_(7), 1145-1156. https://doi.org/10.1111/ele.14226

Parker, T.H., Forstmeier, W., Koricheva, J. et al. (2016). Transparency in Ecology and Evolution: Real problems, real solutions. _Trends in Ecology & Evolution, 31_(9), 711-719. https://doi.org/10.1016/j.tree.2016.07.002

Reid, K., Baker, G.B., & Woehler, E.J. (2023). An ecological risk assessment for the impacts of offshore wind farms on birds in Australia. _Austral Ecology_. https://doi.org/10.1111/aec.13278

Román‐Palacio, C. (2023). The phrutar package: Increasing access, reproducibility and transparency in phylogenetic analyses. _Methods in Ecology and Evolution, 14_(9), 2284-2299. https://doi.org/10.1111/2041-210x.14147

RS-eco (2022). _traitdata: Easy access to various ecological trait data_. R package version 0.0.1, https://github.com/RS-eco/traitdata

Tobias, J.A., Sheard, C., Pigot, A.L. et al. (2022). AVONET: morphological, ecological and geographical data for all birds. _Ecology Letters, 25_(3), 581–597. https://doi.org/10.1111/ele.13898

Westgate, M., Kellie, D., Stevenson, M., Newman, P. (2023). _galah: Biodiversity Data from the Living Atlas Community_. R package version 1.5.3, https://CRAN.R-project.org/package=galah

