---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# dependencies

humdrum relies heavily on tidyverse packages, specifically data.table, dplyr, readxl, and tidyr.

# humdrum

<!-- badges: start -->
<!-- badges: end -->

The goal of humdrum is to interface with the AvoNet dataset and take the humdrum out of exploratory data analysis. Humdrum can install, load, tidy, clean, extract, and summarise AvoNet.

The AvoNet dataset contains functional trait data for 90,020 individual birds representing 11,009 extant species sampled from 181 countries (Tobias et al., 2022). It includes 11 continuous morphological traits (such as beak and wing dimensions; and tail, tarsus, and body size), 6 ecological variables, as well as geographic data. AvoNet also includes species averages according to three different taxonomy classifications to enable integration with other databases.

AvoNet is difficult to explore and interact with as it is stored across 11 related excel spreadsheets. Other datasets stored similarly (e.g. AusTraits, Atlas of Living Australia) have dedicated R-packages to streamline their use (Falster et al., 2021; Westgate et al., 2023). Some of AvoNet is incorporated in the 'traitdata' package, however this (at the time of writing) is only a single observed individual per species according to the BirdLife taxonomy classification (RS-eco, 2022). An R-package that interfaces with the entirety of AvoNet would facilitate EDA and potentially aid in addressing specific research questions.

Humdrum has functions to:
* install and load AvoNet as a relational database (stored as a list of tibbles)
* tidy data when loading (largely correcting variable class and renaming as AvoNet is already in tidy format)
* clean data when loading (toggled based on user preference)
* facilitate Exploratory Data Analysis (EDA) (extract defined variables according to different taxonomies and apply summary statistics)

## Installation

You can install the development version of humdrum from [GitHub](https://github.com/DruryCF/DS4B-final-project) with:

``` r
# install.packages("devtools")
devtools::install_github("DruryCF/DS4B-final-project")
```

## Example

This basic example demonstrates workflow with humdrum:

```{r example}
library(humdrum)

avonet_install(filename = "raw_data/avonet_dataset.xlsx")

avonet <- avonet_load(filename = "raw_data/avonet_dataset.xlsx", clean = "all")

# As the avonet object prints many tibbles, here's some key ones:
# This tibble is built entirely by humdrum to describe what each tibble contains:
avonet$description 

# This is the original Metadata excel spreadsheet with renamed variables:
avonet$metadata

# This is the original AVONET_Raw_Data excel spreadsheet with renamed variables and excluded observations outlined in the avonet_exclude object written to the global environment (also removes empty variables with embargoed data):
avonet$raw_data

# Extract data from $raw_data:
extract <- avonet_extract_intraspecies(filename = "raw_data/avonet_dataset.xlsx",
                            taxon_system = "birdtree",
                            clean = "all")

# Apply summary statistics by species and sex:
avonet_summary(data = extract,
               group = c("species", "sex"),
               stats = c("mean", "median", "max", "min", "sd"),
               tidy = FALSE)
```

# AvoNet

If publishing, be sure to cite the original paper that released AvoNet:

Tobias, J.A., Sheard, C., Pigot, A.L. et al. (2022). AVONET: morphological, ecological and geographical data for all birds. _Ecology Letters, 25_(3), 581–597. https://doi.org/10.1111/ele.13898

# References

Falster, D., Gallagher, R., Wenk, E.H. et al. (2021). AusTraits, a curated plant trait database for the Australian flora. _Sci Data, 8_, 254. https://doi.org/10.1038/s41597-021-01006-6

RS-eco (2022). _traitdata: Easy access to various ecological trait data_. R package version 0.0.1, https://github.com/RS-eco/traitdata

Tobias, J.A., Sheard, C., Pigot, A.L. et al. (2022). AVONET: morphological, ecological and geographical data for all birds. _Ecology Letters, 25_(3), 581–597. https://doi.org/10.1111/ele.13898

Westgate, M., Kellie, D., Stevenson, M., Newman, P. (2023). _galah: Biodiversity Data from the Living Atlas Community_. R package version 1.5.3, https://CRAN.R-project.org/package=galah

Note:
Package was built using the devtools package following https://r-pkgs.org/

Hexsticker was built using the hexSticker package following https://www.youtube.com/watch?v=O34vzdHOaEk
